---
title: "A Quiet Revolution, an Uneven Map, and a Pandemic"
subtitle: "Four decades of childhood vaccination, told in four beats"
author: "Team 15"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 10
    fig-height: 6
    theme: cosmo
    page-layout: article
    code-tools: false
    embed-resources: true
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(viridis)
library(maps)
library(countrycode)
library(ggrepel)
library(ggtext)

vaccine <- read.csv(
  "global-vaccination-coverage.csv",
  check.names = FALSE,
  na.strings  = c("", "NA")
)
gdp <- read.csv(
  "gdp-per-capita-worldbank.csv",
  check.names = FALSE,
  na.strings  = c("", "NA")
) |>
  rename(gdp = `GDP per capita`, region = `World region according to OWID`)

is_country <- function(x) !is.na(x) & nchar(x) == 3 & !grepl("_", x, fixed = TRUE)

countries <- vaccine |> filter(is_country(Code))
world     <- vaccine |> filter(Code == "OWID_WRL")

theme_story <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey35", size = 11,
                                   margin = margin(b = 10)),
      plot.caption  = element_text(color = "grey55", size = 9, hjust = 0),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      axis.title    = element_text(color = "grey25"),
      axis.text     = element_text(color = "grey35"),
      legend.position = "bottom",
      plot.margin   = margin(10, 14, 10, 10)
    )
}
```

::: {.callout-note appearance="simple"}
**Communication goals.** This prototype walks the reader through four
beats: (1) the dramatic global rise in childhood vaccination since 1980,
(2) the inequalities that persist beneath the global average, (3) the
structural — especially economic — drivers of those inequalities, and
(4) how the COVID-19 pandemic disrupted decades of progress. Each beat
is one chart, one paragraph of narration, and one or two annotations
that point directly to the evidence on the chart.
:::

## A quiet revolution, 1980–2024 { #beat-1 }

In 1980, fewer than one in five children worldwide was vaccinated against
diphtheria, tetanus, and pertussis. Today, the figure is more than four
in five. That shift — from **20%** to **85%** in just over four decades —
ranks among the largest public-health gains in modern history, yet it
rarely shows up in headlines. We open the story there: with the long arc
that all of the later complications hang from.

```{r}
#| label: beat1-rise
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Source: WHO / UNICEF Estimates of National Immunization Coverage, via Our World in Data."

world_long <- world |>
  select(Year,
         DTP3 = `Diptheria/tetanus/pertussis (DTP3)`,
         MCV1 = `Measles, first dose (MCV1)`,
         Pol3 = `Polio (Pol3)`,
         HepB3 = `Hepatitis B (HepB3)`) |>
  pivot_longer(-Year, names_to = "vaccine", values_to = "coverage") |>
  filter(!is.na(coverage))

end_pts <- world_long |> group_by(vaccine) |> filter(Year == max(Year)) |>
  ungroup()

ggplot(world_long, aes(Year, coverage, color = vaccine)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 90, ymax = 100,
           fill = "grey85", alpha = 0.35) +
  annotate("text", x = 1981, y = 92.5, label = "WHO 90% target",
           hjust = 0, color = "grey45", size = 3.3, fontface = "italic") +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(data = end_pts, size = 2.6) +
  ggrepel::geom_text_repel(
    data = end_pts,
    aes(label = paste0(vaccine, ": ", coverage, "%")),
    hjust = 0, nudge_x = 1.5, direction = "y",
    size = 3.6, fontface = "bold",
    segment.color = "grey70", segment.size = 0.3,
    box.padding = 0.25, min.segment.length = 0,
    show.legend = FALSE, seed = 1
  ) +
  annotate("segment", x = 1980, xend = 1980, y = 22, yend = 30,
           color = "grey30", linewidth = 0.4) +
  annotate("text", x = 1980, y = 18,
           label = "DTP3 at 20%\nin 1980",
           hjust = 0.05, color = "grey25", size = 3.2, lineheight = 0.9) +
  scale_x_continuous(limits = c(1980, 2030), breaks = seq(1980, 2024, 10)) +
  scale_y_continuous(limits = c(0, 100),
                     labels = function(x) paste0(x, "%")) +
  scale_color_manual(values = c(DTP3 = "#1F4E79", MCV1 = "#C0504D",
                                Pol3 = "#4F81BD", HepB3 = "#9BBB59"),
                     name = NULL) +
  labs(
    title    = "Global childhood vaccination coverage, 1980–2024",
    subtitle = "Share of one-year-olds receiving each vaccine. World-level average.",
    x = NULL, y = NULL,
    caption  = "Shaded band marks the WHO target of 90% coverage."
  ) +
  theme_story() +
  theme(legend.position = "none")
```

Four of the workhorse vaccines tell the same story. Diphtheria-tetanus-pertussis
and polio both crossed the 75% line by 1990 — within a decade of the WHO's
Expanded Programme on Immunization gaining momentum. Hepatitis B took longer
because it was added to the routine schedule later, but it has effectively
caught up. The global picture, in short, looks like a success.

## But "global average" hides a lot { #beat-2 }

A single global line is reassuring, and misleading. Average coverage in 2024
sits at 85%, but the spread *between* countries is still enormous: nearly
**60 percentage points** separate the best- and worst-performing nations.
The map below shows where those gaps live. Two African neighbours,
**Rwanda (98%)** and **Sudan (39%)**, sit at opposite ends of the scale in
the same year.

```{r}
#| label: beat2-map
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5.6
#| fig-cap: "DTP3 coverage among one-year-olds in 2024. Grey = no data reported."

dtp_2024 <- countries |>
  filter(Year == 2024) |>
  transmute(iso3 = Code, Entity, dtp3 = `Diptheria/tetanus/pertussis (DTP3)`)

world_map <- map_data("world")
world_map$iso3 <- suppressWarnings(countrycode(
  world_map$region,
  origin      = "country.name",
  destination = "iso3c",
  warn        = FALSE
))
map_df <- left_join(world_map, dtp_2024 |> select(iso3, dtp3), by = "iso3")

labels <- tribble(
  ~Entity,     ~iso3,  ~lab_x, ~lab_y, ~end_x, ~end_y, ~note,
  "Rwanda",    "RWA",     55,    -18,    30,    -2,   "Rwanda: 98%",
  "Sudan",     "SDN",    -25,     28,    30,    15,   "Sudan: 39%",
  "Azerbaijan","AZE",     85,     55,    48,    40,   "Azerbaijan: 51%\n(GDP > $22K)"
)

ggplot(map_df, aes(long, lat, group = group, fill = dtp3)) +
  geom_polygon(color = "white", linewidth = 0.05) +
  geom_segment(data = labels, inherit.aes = FALSE,
               aes(x = lab_x, y = lab_y, xend = end_x, yend = end_y),
               color = "grey25", linewidth = 0.35) +
  geom_label(data = labels, inherit.aes = FALSE,
             aes(x = lab_x, y = lab_y, label = note),
             size = 3.1, color = "grey15", fill = "white",
             label.size = 0.25, label.padding = unit(0.2, "lines"),
             hjust = 0.5, lineheight = 0.9) +
  coord_fixed(1.3, xlim = c(-170, 180), ylim = c(-58, 85)) +
  scale_fill_viridis_c(
    option   = "viridis",
    name     = "DTP3 coverage",
    limits   = c(0, 100),
    na.value = "grey88",
    breaks   = c(0, 25, 50, 75, 100),
    labels   = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "Vaccination coverage is *uneven* across the world",
    subtitle = "Diphtheria-tetanus-pertussis (DTP3), 2024"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title       = ggtext::element_markdown(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey35", size = 11,
                                    margin = margin(b = 8)),
    legend.position  = "bottom",
    legend.key.width = unit(2.4, "cm"),
    legend.key.height= unit(0.35, "cm")
  )
```

The low-coverage band runs through the Sahel, the Horn of Africa, and a
handful of conflict-affected states; the high-coverage band covers nearly all
of Europe, the Americas, and East Asia. But coverage is not destiny: Rwanda's
98% rate, achieved at a GDP per capita under $3,500, sits next to wealthy
**Azerbaijan's 51%** — a result that should be impossible if vaccination
were simply a question of money.

## Money matters — but it isn't the whole story { #beat-3 }

If wealth alone determined vaccination, every country would line up neatly
along a single curve. They don't. Coverage *does* climb with GDP per capita,
but the cloud of dots is wide enough that nearly every income level contains
both stars and laggards. The trend is real but loose: a Pearson correlation
of roughly **r ≈ 0.5** between log-GDP and DTP3 coverage in 2024.

```{r}
#| label: beat3-gdp
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "GDP per capita on a log scale. Dashed line is a loess fit with 95% CI."

cy <- countries |>
  filter(Year == 2024) |>
  transmute(Code, Entity, dtp3 = `Diptheria/tetanus/pertussis (DTP3)`)

gy <- gdp |>
  filter(Year == 2024, is_country(Code)) |>
  select(Code, gdp, region)

df <- inner_join(cy, gy, by = "Code") |>
  filter(!is.na(dtp3), !is.na(gdp), !is.na(region))

r_val <- cor(log10(df$gdp), df$dtp3)

highlights <- df |> filter(Entity %in% c(
  "Rwanda", "Sudan", "Azerbaijan", "United States",
  "Bangladesh", "Equatorial Guinea"
))

ggplot(df, aes(gdp, dtp3)) +
  geom_smooth(method = "loess", color = "#2C3E50",
              fill = "#2C3E50", alpha = 0.10, linewidth = 0.9,
              linetype = "dashed", se = TRUE, na.rm = TRUE) +
  geom_point(aes(fill = region), shape = 21, color = "white",
             stroke = 0.45, size = 3.4, alpha = 0.9) +
  geom_point(data = highlights, shape = 21, fill = NA,
             color = "black", size = 5, stroke = 1.1) +
  ggrepel::geom_text_repel(
    data = highlights,
    aes(label = Entity),
    size = 3.5, fontface = "bold", color = "grey10",
    box.padding = 0.4, seed = 1
  ) +
  annotate("text", x = 1500, y = 96,
           label = "Punching above its weight\n(Rwanda, Bangladesh)",
           hjust = 0, vjust = 1, size = 3.3, color = "#1B5E20",
           fontface = "italic", lineheight = 0.9) +
  annotate("text", x = 28000, y = 55,
           label = "Wealth without coverage\n(Azerbaijan, Eq. Guinea)",
           hjust = 0.5, size = 3.3, color = "#B71C1C",
           fontface = "italic", lineheight = 0.9) +
  scale_x_log10(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    breaks = c(1000, 2000, 5000, 10000, 20000, 50000, 100000)
  ) +
  scale_y_continuous(limits = c(30, 102),
                     labels = function(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "Africa"        = "#D1495B",
    "Asia"          = "#3E78B2",
    "Europe"        = "#5C9E6E",
    "North America" = "#E08A3C",
    "Oceania"       = "#8E6CA1",
    "South America" = "#4FA8A1"
  ), name = NULL) +
  labs(
    title    = "Coverage rises with wealth — but with a wide cloud",
    subtitle = sprintf(
      "DTP3 coverage vs. GDP per capita, 2024. r = %.2f (log GDP, n = %d).",
      r_val, nrow(df)),
    x = "GDP per capita (log scale, US$)",
    y = NULL
  ) +
  theme_story()
```

The labelled countries are the story. **Rwanda** and **Bangladesh** sit far
above the trend line: poor by world-bank measures, but with delivery systems
strong enough to reach essentially every infant. On the other side,
**Azerbaijan** and **Equatorial Guinea** are oil-rich and yet vaccinate at
rates more typical of much poorer countries. Money helps, but what really
seems to matter is whether a country has a functioning primary-care network
to *use* the money for routine immunization.

## A first decline since 1980 { #beat-4 }

Progress is never guaranteed. Between 2019 and 2021, global DTP3 coverage
fell from **86% to 82%** — the first sustained dip in four decades, driven
by COVID-19 lockdowns and the diversion of healthcare workers to pandemic
response. By 2024 the world had clawed most of it back, but still sits one
percentage point below pre-pandemic peak.

```{r}
#| label: beat4-covid
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5
#| fig-cap: "Four core vaccines, global average. WHO/UNICEF estimates."

recent <- world |>
  select(Year,
         DTP3 = `Diptheria/tetanus/pertussis (DTP3)`,
         MCV1 = `Measles, first dose (MCV1)`,
         Pol3 = `Polio (Pol3)`,
         HepB3 = `Hepatitis B (HepB3)`) |>
  filter(Year >= 2010) |>
  pivot_longer(-Year, names_to = "vaccine", values_to = "coverage")

ggplot(recent, aes(Year, coverage, color = vaccine)) +
  annotate("rect", xmin = 2020, xmax = 2021.5,
           ymin = -Inf, ymax = Inf,
           fill = "#C0504D", alpha = 0.12) +
  annotate("text", x = 2020.75, y = 95,
           label = "COVID-19\ndisruption",
           color = "#7A1F1B", size = 3.4, fontface = "bold",
           lineheight = 0.9) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  ggrepel::geom_text_repel(
    data = recent |> filter(Year == 2024),
    aes(label = paste0(vaccine, " ", coverage, "%")),
    nudge_x = 0.8, direction = "y", hjust = 0,
    size = 3.4, fontface = "bold",
    segment.color = "grey70", segment.size = 0.3,
    box.padding = 0.2, min.segment.length = 0,
    show.legend = FALSE, seed = 1
  ) +
  scale_x_continuous(limits = c(2010, 2025.5),
                     breaks = seq(2010, 2024, 2)) +
  scale_y_continuous(limits = c(78, 96),
                     labels = function(x) paste0(x, "%")) +
  scale_color_manual(values = c(DTP3 = "#1F4E79", MCV1 = "#C0504D",
                                Pol3 = "#4F81BD", HepB3 = "#9BBB59"),
                     name = NULL) +
  labs(
    title    = "COVID-19 broke a 40-year streak — and recovery is incomplete",
    subtitle = "Global coverage of four core vaccines, 2010–2024",
    x = NULL, y = NULL
  ) +
  theme_story()
```

The dip is small in absolute terms but enormous in context: from 1980 through
2019, the line had only ever drifted sideways, never sustained a multi-year
fall. **Measles coverage** has been slowest to recover — and unlike DTP3,
measles needs roughly 95% coverage to prevent outbreaks, so the gap matters.
The dashboard at the end of this report lets you check whether *your* country
has fully recovered.

## What this means — and what to do next

Put together, the four charts tell a single story:

- **Global progress is real.** Childhood vaccination coverage has roughly
  quadrupled in a generation.
- **But the global average hides a 60-point gap** between the best and worst
  performers, and that gap has geographic shape: it concentrates in the
  Sahel and a handful of conflict states.
- **Wealth predicts coverage on average, but not country-by-country.** The
  countries that punch above their weight tend to share a feature wealth
  alone doesn't buy: a working primary-care system.
- **And the system is fragile.** A single global shock cost four years of
  progress, and the recovery isn't yet complete.

Each beat is built directly from the WHO/UNICEF Estimates of National
Immunization Coverage (1980–2024) and World Bank GDP per capita data.

### Now you explore

The static narrative above takes the same dataset and turns it into four
guided moments. The companion dashboard inverts that: it hands the wheel
to the reader. Pick any vaccine, drag through any year, click any country
to see its full trajectory and compare it to the world average.

> **→ Run the dashboard:** `Rscript -e 'shiny::runApp("dashboard")'` from
> the `information_visualization/` folder, or open `dashboard/app.R` in
> RStudio and click **Run App**.

::: {.callout-tip collapse="true"}
## Methods and caveats

- **Population denominator.** Coverage figures describe the share of
  one-year-olds who received each vaccine in that calendar year — not the
  share of the whole population.
- **Missing values are kept as gaps**, not zeros. For newer vaccines
  (PCV3, Rotavirus, IPV1) a blank cell typically means the vaccine wasn't
  yet on the routine schedule.
- **Recent years are partly estimated.** WHO extrapolates from prior
  reporting when countries don't report on time, so 2023–2024 numbers
  should be treated as best-available rather than firm.
- **GDP data starts in 1990.** Any cross-data analyses pre-1990 use only
  the vaccine series.
:::
