---
title: "Three Visualizations of Global Vaccination Coverage"
format:
  html:
    fig-width: 9
    fig-height: 6
    toc: true
author: "Team 15"
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(viridis)
library(maps)
library(countrycode)

vaccine <- read.csv(
  "global-vaccination-coverage.csv",
  sep = ",",
  check.names = FALSE,
  na.strings = c("", "NA")
)

vaccine_cols <- c(
  "Hepatitis B (HepB3)",
  "H. influenza type b (Hib3)",
  "Inactivated polio vaccine (IPV1)",
  "Measles, first dose (MCV1)",
  "Pneumococcal vaccine (PCV3)",
  "Polio (Pol3)",
  "Rubella (RCV1)",
  "Rotavirus (RotaC)",
  "Diptheria/tetanus/pertussis (DTP3)"
)

countries  <- vaccine |> filter(!is.na(Code) & Code != "")
groupings  <- vaccine |> filter(is.na(Code) | Code == "")
```

# 1. Choropleth Map — DTP3 Coverage by Country (2024)

A world map showing the share of one-year-olds who received three doses of the
diphtheria-tetanus-pertussis vaccine in 2024. DTP3 is widely used as a proxy
for the overall health of a country's routine-immunization program, and it is
the most completely reported vaccine in the dataset.

```{r}
#| label: choropleth
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "DTP3 coverage among one-year-olds, 2024. Grey = no data."

dtp_2024 <- countries |>
  filter(Year == 2024) |>
  transmute(iso3 = Code,
            dtp3 = `Diptheria/tetanus/pertussis (DTP3)`)

world_map <- map_data("world")
world_map$iso3 <- countrycode(
  world_map$region,
  origin = "country.name",
  destination = "iso3c",
  warn = FALSE
)

map_df <- left_join(world_map, dtp_2024, by = "iso3")

ggplot(map_df, aes(long, lat, group = group, fill = dtp3)) +
  geom_polygon(color = "white", linewidth = 0.05) +
  coord_fixed(1.3, xlim = c(-170, 180), ylim = c(-58, 85)) +
  scale_fill_viridis_c(
    option = "viridis",
    name = "DTP3 coverage (%)",
    limits = c(0, 100),
    na.value = "grey88",
    breaks = c(0, 25, 50, 75, 100)
  ) +
  labs(
    title = "Global DTP3 vaccination coverage, 2024",
    subtitle = "Share of one-year-olds who received three doses of DTP",
    caption  = "Source: WHO / UNICEF via Our World in Data"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "bottom",
    legend.key.width  = unit(2.2, "cm"),
    legend.key.height = unit(0.35, "cm")
  )
```

**Reading the map.** Most of Europe, the Americas, and East Asia sit at the top
of the scale (~95%+). Persistent low-coverage clusters appear in the Sahel,
parts of the Horn of Africa, and a few conflict-affected states. Greyed-out
regions are countries with no reported value for 2024 (per the documentation,
the most recent two years are partly WHO-estimated).

---

# 2. Multi-Line Time Series — Coverage Trajectories by Region, 1980–2024

How has coverage changed over 45 years? Here we use the grouping rows in the
dataset (the six UNICEF regions plus the global `World` series) and plot each
vaccine as its own small multiple. NA values are left as gaps rather than
zero, because for newer vaccines a blank cell means "not yet in the routine
schedule," not "0% coverage."

```{r}
#| label: timeseries
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "Coverage over time for each vaccine, by UNICEF region and globally."

region_long <- groupings |>
  filter(Entity != "Middle-income countries") |>
  pivot_longer(
    cols = all_of(vaccine_cols),
    names_to  = "vaccine",
    values_to = "coverage"
  ) |>
  mutate(
    Entity = gsub(" \\(UNICEF\\)", "", Entity),
    vaccine = factor(vaccine, levels = vaccine_cols)
  )

ggplot(region_long, aes(Year, coverage, color = Entity)) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  facet_wrap(~ vaccine, ncol = 3) +
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)) +
  scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  scale_color_brewer(palette = "Dark2", name = NULL) +
  labs(
    title = "Vaccination coverage over time, by region",
    subtitle = "Share of one-year-olds vaccinated, 1980–2024",
    x = NULL, y = "Coverage (%)",
    caption  = "Source: WHO / UNICEF via Our World in Data"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey30"),
    strip.text      = element_text(face = "bold", size = 9),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

**Reading the chart.** A few stories jump out:

- The classic 1980s ramp-up is visible for **DTP3**, **Polio**, **Measles
  (MCV1)**, and **Rubella** — these are the vaccines whose facets stretch all
  the way back to 1980.
- **PCV3**, **Rotavirus**, and **IPV1** appear only in recent years because
  they were added to the routine schedule later (2008, 2006, and 2015
  respectively).
- The visible 2020–2021 dip across nearly every panel is the COVID-19
  pandemic's effect on routine immunization services.
- **West and Central Africa** and **Eastern and Southern Africa** consistently
  trail the other regions; **East Asia and the Pacific** and **Latin America**
  sit near the global average or above.

---

# 3. Correlation Heatmap — How Co-Administered Are the Nine Vaccines?

Pairwise Pearson correlations between the nine vaccine columns, computed
across all country-year observations and reordered by hierarchical clustering
so that similar vaccines sit next to each other.

```{r}
#| label: correlation
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 7
#| fig-cap: "Pairwise Pearson correlation between vaccines (country-year level)."

cor_mat <- countries |>
  select(all_of(vaccine_cols)) |>
  cor(use = "pairwise.complete.obs")

ord <- hclust(as.dist(1 - cor_mat))$order
cor_mat <- cor_mat[ord, ord]

cor_df <- as.data.frame(as.table(cor_mat)) |>
  rename(vaccine_x = Var1, vaccine_y = Var2, r = Freq)

ggplot(cor_df, aes(vaccine_x, vaccine_y, fill = r)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r)),
            color = "grey15", size = 3.4) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limits = c(-1, 1),
    name = "Pearson r"
  ) +
  coord_fixed() +
  labs(
    title = "How correlated are vaccine coverage rates?",
    subtitle = "Across all country-year observations, clustered ordering",
    x = NULL, y = NULL,
    caption  = "Source: WHO / UNICEF via Our World in Data"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30"),
    axis.text.x   = element_text(angle = 35, hjust = 1),
    panel.grid    = element_blank(),
    legend.position = "right"
  )
```

**Reading the heatmap.** Clustering pulls out two clear groups:

- A tight **"legacy bundle"** of DTP3, Polio, HepB, and Measles, all correlated
  at $r > 0.85$. These vaccines have shared delivery infrastructure for
  decades, so countries that do well on one almost always do well on the
  others.
- A looser cluster of the **newer vaccines** — PCV3, Rotavirus, and IPV1 —
  which correlate strongly with each other (they were rolled out into the same
  countries around the same time) but slightly less strongly with the legacy
  bundle.
- **Rubella (RCV1)** sits between the two, which fits the fact that it is
  typically delivered as the combined measles–rubella vaccine.
